---
title             : "NFL 2015: A Statistical Analysis"
# shorttitle        : "A Statistical Analysis"

author: 
  - name          : "Dave Dyer"
    affiliation   : "1"
  - name          : "Jostein Barry-Straume"
    affiliation   : "1"
  - name          : "Robert Flamenbaum"
    affiliation   : "1" 
  - name          : "Bryan Cikatz"
    affiliation   : "1"

affiliation:
  - id            : "1"
    institution   : "Southern Methodist University"

# abstract: |  



  
# keywords          : "keywords"
# wordcount         : "X"

bibliography      : ["r-references.bib"]

figsintext        : yes
figurelist        : no
tablelist         : no
footnotelist      : no
lineno            : no

lang              : "english"
class             : "jou"
output            : papaja::apa6_pdf
---

```{r load_packages, include = FALSE}
library("papaja")
```

```{r analysis_preferences}
# Seed for random number generation
set.seed(42)
```

# Abstract
 This study investigates how average and variance of yards gained impacts a given team's ability to get touchdowns in the NFL. The theory is that teams with high average yards gained, but with low variance, would get more touchdowns than those with higher average yards gained, but higher variance.  In other words, consistency is more important than big plays.

# The Data
 The data are the 2015 play-by-play records available at Kaggle (https://www.kaggle.com/maxhorowitz/nflplaybyplay2015).  These data are comprised of 46,129 rows describing every NFL play run over the 2015 season, by 32 teams.  There are 65 columns that include multiple values, including text description of the play / penalty, names of players , touchdown boolean flags, down markers, timestamps, game ids, and yardages.

# The Tools
For this analysis, we use MongoDB for data mining and basic statistics, and R for data visualization and Markdown.  We originally had planned on using MySQL on bluemix, but abandoned it in favor of Mongo after some serious issues whist loading the data. 

# EDA
We counted the number of plays per team.  Here's that breakdown:

\newpage

```{r table, echo=FALSE}
library(knitr)
nflwins  <- data.frame(
  teams = c("broncos", "chargers", "oilers"),
  wins = c(10, 12, 5), 
  losses = c(4, 2, 9)
)

kable(nflwins, caption= "W/L Data")
```


# Results

# Discussion
We used `r cite_r("r-references.bib")` for all our analyses.


# Conclusion

# Appendices


\newpage

# References
```{r create_r-references}
r_refs(file = "r-references.bib")
```

\setlength{\parindent}{-0.5in}
\setlength{\leftskip}{0.5in}
