---
title             : "NFL 2015: A Statistical Analysis"
author            : "Dave Dyer, Jostein Barry-Straume, Bryan Cikatz and Robert Flamenbaum"
date              : "November 28, 2017"
output: 
  pdf_document: 
    pandoc_args: [
      "-V", "classoption=twocolumn"
      ]
---


```{r load_packages, include = FALSE}
#library("papaja")
library("mongolite")
library("dplyr") 
library("knitr")
library("tibble")
library("kableExtra")
jostein_uri <- "mongodb://admin:KFTZXKMBGIWKMWJK@sl-us-south-1-portal.12.dblayer.com:28046,sl-us-south-1-portal.17.dblayer.com:28046/admin?3t.connection.name=JosteinDB&3t.connectTimeout=10000&3t.uriVersion=2&3t.certificatePreference=RootCACert:accept_any&3t.databases=admin,compose&3t.useClientCertPassword=false&readPreference=primary&ssl=true&3t.socketTimeout=0&3t.sharded=true&3t.connectionMode=multi"
m <- mongo(collection = "NFL", db = "test",  verbose = TRUE, url=jostein_uri, options = ssl_options(weak_cert_validation = T))
m$count('{}')
alldata <- m$find('{}')
head(alldata)
```

# Abstract
 This study investigates how average and variance of yards gained impacts a given team's ability to get touchdowns in the NFL. The theory is that teams with high average yards gained, but with low variance, would get more touchdowns than those with higher average yards gained, but higher variance.  In other words, consistency is more important than big plays.

# Related Work
There are two related works that also use NFL play-by-play data to ascertain trends in yardage and touchdowns.  The first, "Underrated NFL Stats" [@Iyer] looks into the big plays allowed by the defense and the teams with the best / worst big plays allowed stats.  The second, "How to Quantify the NFL" [@Kelly] looks at big play differentials, but doesn't actually do any statistical analysis.  Our analysis focuses on the variance of the yardage per play numbers and measures the efficacy of teams by measuring the touchdowns from each game & team.


# Introduction
The data are the 2015 play-by-play records available at Kaggle (https://www.kaggle.com/maxhorowitz/nflplaybyplay2015).  These data are comprised of 46,129 rows describing every NFL play run over the 2015 season, by 32 teams.  There are 65 columns that include multiple values, including text description of the play / penalty, names of players , touchdown boolean flags, down markers, timestamps, game ids, and yardages.  For the purposes of this study, we focused on the average and variance of yards per play, by team, and measured the 'success' of the team based on the count of touchdowns per game.  

```{r mongolite_queries, include=FALSE}
# Returns all plays that scored a touchdown with positive yard gains,
# grouping by team and count the plays per team, sorted in alphabetical order
ydsGainTD <- m$aggregate('[
		{ "$match": { "Yards.Gained": { "$gt": 0 }, "Touchdown": { "$gt": 0 } } },
    { "$group": { "_id": "$posteam" , "count": { "$sum": 1 } } },
    { "$sort": { "_id": 1} }
  ]')
ydsGainTD.tibble <- ydsGainTD

# Returns average yards gained and average number of touchdowns per team
avgYdsTds <- m$aggregate('[
  { "$group": {
      "_id": "$posteam",
      "ydsGainAvg": { "$avg": "$Yards.Gained" },
      "ydsGainSd": { "$stdDevPop": "$Yards.Gained"},
      "ydsGainSdSample": { "$stdDevSamp": "$Yards.Gained"},
      "tdsAvg": { "$avg": "$Touchdown" },
      "tdsSum": { "$sum": "$Touchdown" }
      }
  }]')
avgYdsTds.tibble <- avgYdsTds$`_id`

# Returns sum of touchdowns and sum of penalty yards per team
sumTdsPenYds <- m$aggregate('[
  { "$group": { 
      "_id": { "posteam": "$posteam"},
      "Touchdown": { "$sum": "$Touchdown" },
      "Penalty_Yards": { "$sum": "$Penalty.Yards" }
    }
  }]')
sumTdsPenYds.tibble <- sumTdsPenYds$`_id`

# Returns sum of touchdowns by quarter per team
sumTdsPerQtr <- m$aggregate('[
  { "$group": { 
    "_id": {"Team": "$posteam", "Quarter": "$qtr"},
    "Touchdown": { "$sum": "$Touchdown" }
    }
  }]')
sumTdsPerQtr.tibble <- sumTdsPerQtr$`_id`

ydsGainVTd <- m$aggregate('[
  { "$group": {
      "_id": { "GameID": "$GameID", "PosTeam":"$posteam"},
      "SumTDs": { "$sum": "$Touchdown"},
      "SumYdsGain": { "$sum": "$Yards.Gained"}, 
      "AvgYdsGain": { "$avg": "Yards.Gained"},
      "SDYdsGain": { "$stdDevPop": "$Yards.Gained"}
  }}]')
ydsGainVTd <- ydsGainVTd[ydsGainVTd$SumTDs!=0,]
```

# The Tools
For this analysis, we use MongoDB for data mining and basic statistics, and R for data visualization and Markdown.  We originally had planned on using MySQL on bluemix, but abandoned it in favor of Mongo after some serious issues whist loading the data.  

We used Mongodb hosted on IBM Bluemix and Rstudio with Rmarkdown for the analysis, visualization and writeup.  We used many r packages to support our analysis and data mining.  For data mining, we used the mongolite package (https://jeroen.github.io/mongolite/) to connect to the bluemix data set and, where possible, we used mongo to do the statistical analysis in favor of R.  Below are the specifications for the Mongo Database:  
  
  

```{r tools, echo = FALSE}
words <- c("Bluemix Storage", "Data Size", "Database Server", "Database Version", "Database Location", "Cloud Hosting Service", "Processors", "Memory")
numbers <- c("1GB", "14.3MB", "Compose for MongoDB-jj", "3.2.11", "US South", "IBM Bluemix", "1 x 2.0 GHz Cores", "1GB RAM")
specs <- data.frame( "Parameter"=words,"Specification"= numbers)
kable(specs, format="latex") %>% kable_styling(latex_options="striped", font_size=8)
```

# Analysis

When the spread of per-team yards-per-play is viewed in a team-specific box \& whisker plot, it is easy to see that there is not a great amount of variance between the medians; most teams are between 5 and 6 median yards per play, with handful of teams above and below this range.  


```{r figure1, echo = FALSE}
library(ggplot2)
bygame <- alldata %>%
  group_by_at(vars(posteam, GameID)) %>%
  summarize(TDgameSum = sum(Touchdown),
            avgYdsPerGamePerTeam = mean(Yards$Gained),
            sdYdsPerGamePerTeam = sd(Yards$Gained))
bygame <- bygame[bygame$avgYdsPerGamePerTeam!=0,]
bp1 <- ggplot(data = bygame, aes(posteam, x = posteam, y = avgYdsPerGamePerTeam))
bp1 + geom_boxplot() +
  ylab("Yards Per Play") + 
  xlab("Team") + 
  ggtitle("Median Yards per Play (by Team)") + 
  geom_hline(yintercept = median(bygame$avgYdsPerGamePerTeam), color = "blue") +  # Added horizontal line for the median of avgYdsPerGamePerTeam
  scale_y_continuous(limits = c(0, 10))
```  

This is about what we'd expect, since the NFL is incredibly competitive, and self-normalizing due to constant trading, strategy, talent balance, and strategic positional matchups that, on balance, make the league pretty competitive.  

If one team were overl dominant in median yards per play, we would expect that team to dominate the entire season, year after year, which is not the case.  However, there *is* quite a bit of difference in the outer quartiles and in the number of outliers.  These data support our investigation into whether or not a large variance changes team performance.

## Yards v. Touchdowns 
To start, we decided to plot number of touchdowns against yards gained (per game).  We expect to see a linear trend upwards that illustrates the basic football concept of more yards = more points.  As you can see, there appears to be an upward trend -- that is, as yards increase, so does the number of touchdowns.  

```{r figure2, echo = FALSE, fig.pos="H"}
# Calculate intercept and slope for AB Line
coefs2 <- coef(lm(SumYdsGain ~ SumTDs, data = ydsGainVTd))
p2 <- ggplot(ydsGainVTd, aes(SumTDs, SumYdsGain)) 
p2 + geom_jitter() + 
  xlab("Touchdowns") + 
  ylab("Yards") + 
  ggtitle("Touchdowns vs. Yards (Per Game)") +
  geom_abline(intercept = coefs2[1], slope = coefs2[2], color = "blue") + 
  scale_x_continuous(limits=c(0,9)) + 
  scale_y_continuous(limits=c(0, 800))
```  

However, the theory of this paper is that, if yards per play is normalized, the lower variance teams will get more touchdowns per game.  So it is important to know how the non-normalized standard deviation of yards per play relates to touchdowns per game on average.

## Std Dev v. Touchdowns
The below plot shows that it is relatively static, with perhaps a slight rise as standard deviations go up.  


```{r figure3, echo = FALSE, fig.pos="H"}
# Calculate intercept and slope for AB Line
coefs3 <- coef(lm(SDYdsGain ~ SumTDs, data = ydsGainVTd))
ggplot(ydsGainVTd, aes(SumTDs, SDYdsGain, group = "_id.GameID")) + 
  geom_jitter() + 
  xlab("Touchdowns") + 
  ylab("StDev in Yards (Per Game)") + 
  ggtitle("Touchdowns vs. Variance in Yards Per Game") + 
  geom_abline(intercept = coefs3[1], slope = coefs3[2], color = "blue") + 
  scale_x_continuous(limits=c(0,9)) + 
  scale_y_continuous(limits=c(0, 17))
```

However, the data are not normalized, and in order to understand the impact a normalization routine would have on the data, we can examine the relationship between the standard deviation of yards per play and the total yards per game.  We expect that, as yards per game goes up, so does total yards, but that is not what's interesting about this plot.  Note that it's not uniform linear growth as the standard deviation and the yards per game increase.  The spread increases as yards and standard deviation increase, indicating that you get more yards with more variance, but you also run the risk of not getting as many yards (presumably because you're taking more chances, and you get higher rewards for higher risk... sometimes.)  


```{r figure4, echo = FALSE, fig.pos="H"}
# Calculate intercept and slope for AB Line
coefs4 <- coef(lm(SumYdsGain ~ SDYdsGain, data = ydsGainVTd))
ggplot(ydsGainVTd, aes(SDYdsGain, SumYdsGain, group = "_id.GameID")) + 
  geom_jitter() + 
  ylab("Yards") + 
  xlab("StDev(Yards)") + 
  ggtitle("Yards vs. StDev(Yards) per Game") + 
  geom_abline(intercept = coefs4[1], slope = coefs4[2], color="blue") + 
  scale_x_continuous(limits=c(0,17)) + 
  scale_y_continuous(limits=c(0, 800))
```
 


# Conclusion

The positive slope of the fit line in the StDev vs. Yards per Game plot is indicative that the surprising result in figure 3 -- that is, that higher standard deviation equals more touchdowns -- could be explained by this strong increasing trend.  In order to find out for sure, we need a way to remove the impact of Yards / Game on Touchdowns / Game.  Unfortunately, such an exercise is outside of the scope of this class.

# Appendices
Below are the R code chunks used for this markdown.  

```{r code, eval=FALSE, echo=TRUE, message=FALSE}
# Returns all plays that scored a touchdown with positive yard gains,
# grouping by team and count the plays per team, sorted in alphabetical order
ydsGainTD <- m$aggregate('[
		{ "$match": { "Yards.Gained": { "$gt": 0 }, "Touchdown": { "$gt": 0 } } },
    { "$group": { "_id": "$posteam" , "count": { "$sum": 1 } } },
    { "$sort": { "_id": 1} }
  ]')
ydsGainTD.tibble <- ydsGainTD

# Returns average yards gained and average number of touchdowns per team
avgYdsTds <- m$aggregate('[
  { "$group": {
      "_id": "$posteam",
      "ydsGainAvg": { "$avg": "$Yards.Gained" },
      "ydsGainSd": { "$stdDevPop": "$Yards.Gained"},
      "ydsGainSdSample": { "$stdDevSamp": "$Yards.Gained"},
      "tdsAvg": { "$avg": "$Touchdown" },
      "tdsSum": { "$sum": "$Touchdown" }
      }
  }]')
avgYdsTds.tibble <- avgYdsTds$`_id`

# Returns sum of touchdowns and sum of penalty yards per team
sumTdsPenYds <- m$aggregate('[
  { "$group": { 
      "_id": { "posteam": "$posteam"},
      "Touchdown": { "$sum": "$Touchdown" },
      "Penalty_Yards": { "$sum": "$Penalty.Yards" }
    }
  }]')
sumTdsPenYds.tibble <- sumTdsPenYds$`_id`

# Returns sum of touchdowns by quarter per team
sumTdsPerQtr <- m$aggregate('[
  { "$group": { 
    "_id": {"Team": "$posteam", "Quarter": "$qtr"},
    "Touchdown": { "$sum": "$Touchdown" }
    }
  }]')
sumTdsPerQtr.tibble <- sumTdsPerQtr$`_id`

ydsGainVTd <- m$aggregate('[
  { "$group": {
      "_id": { "GameID": "$GameID", "PosTeam":"$posteam"},
      "SumTDs": { "$sum": "$Touchdown"},
      "SumYdsGain": { "$sum": "$Yards.Gained"}, 
      "AvgYdsGain": { "$avg": "Yards.Gained"},
      "SDYdsGain": { "$stdDevPop": "$Yards.Gained"}
  }}]')
ydsGainVTd <- ydsGainVTd[ydsGainVTd$SumTDs!=0,]

library(ggplot2)
bygame <- alldata %>%
  group_by_at(vars(posteam, GameID)) %>%
  summarize(TDgameSum = sum(Touchdown),
            avgYdsPerGamePerTeam = mean(Yards$Gained),
            sdYdsPerGamePerTeam = sd(Yards$Gained))
bp1 <- ggplot(data = bygame, 
              aes(posteam, 
                  x = posteam,
                  y = avgYdsPerGamePerTeam))
bp1 + geom_boxplot() + ylab("Yards Per Play") + xlab("Team") + 
ggtitle("Median Yards per Play (by Team)") + 
geom_hline(yintercept = median(bygame$avgYdsPerGamePerTeam), color = "blue") 
# Added horizontal line for the median of avgYdsPerGamePerTeam

# Calculate intercept and slope for AB Line
coefs2 <- coef(lm(SumYdsGain ~ SumTDs, data = ydsGainVTd))
p2 <- ggplot(ydsGainVTd, aes(SumTDs, SumYdsGain)) 
p2 + geom_jitter() + 
  xlab("Touchdowns") + 
  ylab("Yards") + 
  ggtitle("Touchdowns vs. Yards (Per Game)") +
  geom_abline(intercept = coefs2[1], slope = coefs2[2], color = "blue") + 
  scale_x_continuous(limits=c(0,9)) + 
  scale_y_continuous(limits=c(0, 800))

# Calculate intercept and slope for AB Line
coefs3 <- coef(lm(SDYdsGain ~ SumTDs, data = ydsGainVTd))
ggplot(ydsGainVTd, aes(SumTDs, SDYdsGain, group = "_id.GameID")) + 
  geom_jitter() + 
  xlab("Touchdowns") + 
  ylab("StDev in Yards (Per Game)") + 
  ggtitle("Touchdowns vs. Variance in Yards Per Game") + 
  geom_abline(intercept = coefs3[1], slope = coefs3[2], color = "blue") + 
  scale_x_continuous(limits=c(0,9)) + 
  scale_y_continuous(limits=c(0, 17))

# Calculate intercept and slope for AB Line
coefs4 <- coef(lm(SumYdsGain ~ SDYdsGain, data = ydsGainVTd))
ggplot(ydsGainVTd, aes(SDYdsGain, SumYdsGain, group = "_id.GameID")) + 
  geom_jitter() + 
  ylab("Yards") + 
  xlab("StDev(Yards)") + 
  ggtitle("Yards vs. StDev(Yards) per Game") + 
  geom_abline(intercept = coefs4[1], slope = coefs4[2], color="blue") + 
  scale_x_continuous(limits=c(0,17)) + 
  scale_y_continuous(limits=c(0, 800))
```

# References
Insert references here.
```{r create_r-references}
#r_refs(file = "r-references.bib")
```

\setlength{\parindent}{-0.5in}
\setlength{\leftskip}{0.5in}
